---
title: "MiniProject1"
output: html_document
---

```{r, echo=FALSE}
library(bsts)
library(bpa)
```

The SST data set describes monthly sea temperature sampled every 10 meters from 10 to 90 meters. It has **158 records** ranging from **January 2004 to April 2018**.

Data pattern/grouping? (bpa package)

### Sea Surface Temperature Dataset
Import the dataset from github
Convert the string date into a date type 
Realign the columns so the time features come first then the various depth lengths
training and test sets: 126 : 32
```{r, echo=FALSE}
sea <- read.csv("https://github.com/asbates/bayes-time-series/raw/master/data/gilbralter_time_series_r_2.csv", header=TRUE)
sea$startDate <- as.Date(sea$startDate)
sea <- sea[, c(1, 3, 2, 4, 5, 6, 7, 8, 9, 10, 11, 12)]

new_names <- c("time_index", "date", "SST", "10m", "20m", "30m", "40m", "50m", "60m", "70m", "80m", "90m")

names(sea) <- new_names

head(sea, n=6)

train = sea[1:126, ] ; test = sea[127: 158, ]

```


```{r}

names(sea)[1:length(sea)]
```

Insights 
```{r, echo=FALSE}
#print(paste("rows: ", nrow(sea), "columns: ", ncol(sea)))

#print(min(sea$startDate)) ; print(max(sea$startDate))

```

Insights 
### Density Plots 
Here are density plots for surface levels 10-90
```{r, echo=FALSE}
# density plots of the sea depths 0 meters - 90 meters
par(mfrow=c(3, 2))
#par(mfrow=c(4, 3))


#x0 <- sea$`10m`
#hist(x0, main = "Surface Level Sea Temperature", xlab="Temperature", breaks=20, prob=TRUE)
#lines(density(x0))
#abline(v=mean(x0), col="blue", lwd=2)


x10 <- sea$`10m`
hist(x10, main = "Sea Temperature at 10 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x10))
abline(v=mean(x10), col="blue", lwd=2)


x20 <- sea$`20m`
hist(x20, main = "Sea Temperature at 20 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x20))
abline(v=mean(x20), col="blue", lwd=2)


x30 <- sea$`30m`
hist(x30, main = "Sea Temperature at 30 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x30))
abline(v=mean(x30), col="blue", lwd=2)


x40 <- sea$`40m`
hist(x40, main = "Sea Temperature at 40 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x40))
abline(v=mean(x40), col="blue", lwd=2)


x50 <- sea$`50m`
hist(x50, main = "Sea Temperature at 50 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x50))
abline(v=mean(x50), col="blue", lwd=2)

par(mfrow=c(3, 2))

x60 <- sea$`60m`
hist(x60, main = "Sea Temperature at 60 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x60))
abline(v=mean(x60), col="blue", lwd=2)


x70 <- sea$`70m`
hist(x70, main = "Sea Temperature at 70 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x70))
abline(v=mean(x70), col="blue", lwd=2)


x80 <- sea$`80m`
hist(x80, main = "Sea Temperature at 80 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x80))
abline(v=mean(x80), col="blue", lwd=2)


x90 <- sea$`90m`
hist(x90, main = "Sea Temperature at 90 Meters", xlab="Temperature", breaks=20, prob=TRUE)
lines(density(x90))
abline(v=mean(x90), col="blue", lwd=2)
```

Insights
### Line Graphs x0-x90

We see strong seasonality across all depths but overall temperatures are lower the deeper the depth

```{r, echo=FALSE}
par(mfrow=c(3, 2))

#plot(sea$startDate, sea$X0, main ="Sea Temperature: Surface Level", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1) 

#plot(sea$startDate, sea$X0, main ="Sea Temperature: Surface Level", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1) 

plot(sea$date, sea$`10m` , main ="Sea Temperature: 10 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`20m`, main ="Sea Temperature: 20 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`30m`, main ="Sea Temperature: 30 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`40m`, main ="Sea Temperature: 40 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`50m`, main ="Sea Temperature: 50 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`60m`, main ="Sea Temperature: 60 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`70m`, main ="Sea Temperature: 70 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`80m`, main ="Sea Temperature: 80 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

plot(sea$date, sea$`90m`, main ="Sea Temperature: 90 Meters", xlab="Time", ylab="Temperature", ylim=c(15, 25), type = "l", lty=1)

```

Will introducing covariates improved forecasting?
- sea surface temperature near gibraltar
- aggregated monthly
- January 2004 - November 2017
- Covariates: 10 meter thick water layers at 10 - 90 meters

we are looking at 10 different time series, each separated by 10 meters of water

Regression
- y is temperature (output)
- x are the layers (input)

### Bayesian Modelling
```{r echo=FALSE}
# this section fits one local linear, seasonal, and one linear component
# ping sets a print statement while the function runs
# expected model size sets our spike & slab to have one spike
  # we are expected 1 depth to influence temperature underneath the former mentioned layer
nseasons <- 11
ss <- list()
ss <- AddLocalLinearTrend(ss, y=sea$SST)
ss <- AddSeasonal(ss, sea$SST, nseasons = nseasons)
rlls_model = bsts(SST ~`10m`+`20m`+`30m`+`40m`+`50m`+`60m`+`70m`+`80m`+`90m`, state.specification = ss, data=train, niter=1000, ping=0, expected.model.size=1)


```

### Component Plotting & Coefficient Significance
the 10-20 depth has a high inclusion probability 
white bars are positive betas & black bars have negative betas
```{r, echo=FALSE}
plot(rlls_model, "components", xlab="time_index [ ]", ylab= "SST [deg C]")

# plot coefficients for significance
  # look at  
plot(rlls_model, "coefficients")
```

### Forecasting Model 1
With our prediction on the test set we notice wide confidence intervals. We suspect this is due to the stationary nature of BSTS.
```{r, echo=FALSE}
rlls_model_pred <- predict(rlls_model, newdata = test, horizon= 28)

plot(rlls_model_pred, plot.original=90,
     main="Seas. Local Lin. Trend Forecast With Regression", xlab="time_index [ ]", ylab= "SST [deg C]")


```


### MSPE Model 1
```{r}
errors = bsts.prediction.errors(rlls_model, cutpoints = 9, burn = SuggestBurn(.1, rlls_model), standardize = FALSE)


PlotDynamicDistribution(errors$in.sample)

summary(rlls_model, burn = SuggestBurn(.1, rlls_model))

```


### Model 2 Local Linear, Seasonal, TWO Linear Component Model
This model checks the belief that there are two variables that affect sea surface temperature
```{r}
r2lls_model <- bsts(SST ~`10m`+`20m`+`30m`+`40m`+`50m`+`60m`+`70m`+`80m`+`90m`, state.specification = ss, data=train, niter=1000, ping=0, expected.model.size=2)

#plot(r2lls_model, "coefficients")


```

### Forecasting (Model 2)
```{r}
r2lls_model_pred <- predict(r2lls_model, newdata = test, horizon= 28)

plot(r2lls_model_pred, plot.original=90,
     main="BSTS With Regression", xlab="time_index [ ]", ylab= "SST [deg C]")





```

### MSPE (Model 2)
```{r}
errors2 = bsts.prediction.errors(r2lls_model, cutpoints = NULL, burn = SuggestBurn(.1, rlls_model), standardize = FALSE)


PlotDynamicDistribution(errors2$in.sample)

summary2 <- summary(r2lls_model, burn = SuggestBurn(.1, rlls_model))

```



### Model 3
Specifying inclusion probabilities
```{r}
bp <- c(0.6, 0.3, 0.3, 0.3, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1)
bp_model <- bsts(SST ~`10m`+`20m`+`30m`+`40m`+`50m`+`60m`+`70m`+`80m`+`90m`, state.specification = ss, data=train, niter=1000, ping=0, expected.model.size=2,
                 prior.inclusion.probabilities=bp)

```

### Model Comparison (Model 1, 2, 3)
```{r}
CompareBstsModels(lwd = 2, model.list = list("Model 1"=rlls_model, "Model 2"=r2lls_model,
                                             "Model 3"= bp_model), xlab = "time_index []")


```

Experimental code following class lecture example. 
```{r}
data(sea)
y <- log(sea$SST)
ss <- AddLocalLinearTrend(list(), y)
ss <- AddSeasonal(ss, y, nseasons = 11)
model <- bsts(y, state.specification = ss, niter = 1000)
pred <- predict(model, mewdata= test ,horizon = 28, burn = 100)
plot(pred, plot.original=90)

```
